```{r setup}
#| include: false
# rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(tidyverse)
library(terra)
library(sf)
library(patchwork)
path <- "~/Documents/data/straore/"
```

# S. Traore {.unnumbered}

*Description.*

## Biomass

```{r biomass}
#| eval: false
save_path <- file.path(path, "agb_change")
dir.create(save_path)

biomass <- list.files(file.path(path, "agb_map_masked"), 
           pattern = "map_with_veg", full.names = TRUE) %>% 
  rast() %>% 
  project("EPSG:32630")

biomass <- biomass * (res(biomass)[1]^2 / 10^4)

names(biomass) <- gsub("map_with_veg_", "", names(biomass))
values(biomass)[is.na(values(biomass))] <- 0

years <- as.numeric(names(biomass)[-1])
stats <- data.frame(year = years,
                    deforest_agb = NA,
                    loss_agb = NA,
                    gain_agb = NA,
                    deforest_area = NA,
                    loss_area = NA,
                    gain_area = NA)

seuil <- 0.01

for(y in years){
  biomass0 <- biomass[[as.character(y-1)]]
  biomass1 <- biomass[[as.character(y)]]
  
  # delta AGB & no_forest
  delta <- biomass1 - biomass0    
  
  # deforestation: delta AGB < 0 & no_forest (biomass1 == 0)
  deforest <- delta * (biomass1 == 0) * (biomass0 > 0)
  
  # loss: delta AGB < 0 & forest (biomass1 > 0)
  loss <- delta * (biomass1 > 0) * (biomass0 > 0) * (delta < 0)
  
  # filtered loss
  floss <- loss
  floss[-loss < seuil*biomass0] <- 0
  
  # gain: delta AGB < 0 & forest (biomass1 > 0)
  gain <- delta * (biomass1 > 0) * (biomass0 > 0)  * (delta > 0)
  
  # save stat
  fact <- res(deforest)[1]^2/10^4
  stats[stats$year == y,2:7] <- c(sum(values(deforest)), 
                                  sum(values(loss)),
                                  sum(values(gain)),
                                  -sum(values(deforest < 0))*fact,
                                  -sum(values(-loss > seuil*biomass0))*fact,
                                  sum(values(gain > seuil*biomass0))*fact)
  
  # raster
  writeRaster(deforest, overwrite = T,
          filename = file.path(save_path, 
                               paste0("deforest_", y, ".tif")))
  writeRaster(loss, overwrite = T,
              filename = file.path(save_path, 
                                   paste0("loss_", y, ".tif")))
    writeRaster(floss, overwrite = T,
              filename = file.path(save_path, 
                                   paste0("floss_", y, ".tif")))
  writeRaster(loss/biomass0*100, overwrite = T,
              filename = file.path(save_path, 
                                   paste0("ploss_", y, ".tif")))
  writeRaster(gain, overwrite = T,
              filename = file.path(save_path, 
                                   paste0("gain_", y, ".tif")))
}

stats <- stats %>% 
  mutate(change_agb = gain_agb + loss_agb) %>% 
  mutate(change_area = gain_area + loss_area) %>% 
  gather(variable, value, -year) %>% 
  group_by(variable) %>% 
  arrange(year) %>% 
  mutate(cumul = cumsum(value)) %>% 
  separate(variable, c("type", "variable"))
  
write_tsv(stats, file = file.path(path, "stats.tsv"))
```

```{r}
#| message: false
read_tsv(file = file.path(path, "stats.tsv")) %>% 
  filter(type %in% c("change", "deforest")) %>% 
  mutate(variable_long = recode(variable, 
                                "agb" = "Biomasse (MtC)",
                                "area" = "Surface (Mha)")) %>% 
  mutate(type_long = recode(type,
                            "change" = "Degradation",
                            "deforest" = "Deforestation")) %>% 
  ggplot(aes(year, cumul/10^6, col = type_long)) +
  geom_line() +
  theme_bw() +
  geom_smooth(se = FALSE) +
  facet_wrap(~ variable_long, scales = "free") + 
  scale_color_discrete("") +
  theme(axis.title = element_blank(), legend.position = "bottom")
```

```{r}
ploss <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "ploss", full = TRUE) %>% 
  rast() %>% 
  as.data.frame() %>% 
  gather(year, loss) %>% 
  filter(loss < 0)
data <- sample(log10(-ploss$loss), 10^5)
```

```{r}
#| eval: false
library(VGAM)
fit <- vglm(data ~ 1, 
            mix2normal(eq.sd = FALSE), 
            iphi=0.005, imu=0, imu2=2, isd1=1, isd2=1)
pars = as.vector(coef(fit))
w = logit(pars[1], inverse=TRUE)
m1 = pars[2]
sd1 = exp(pars[3])
m2 = pars[4]
sd2 = exp(pars[5])
```

-   poids: 0.3483403
-   mu1: -0.03484159, soit 0.92%
-   sd1: 0.6321144
-   mu2: 0.7023864, soit 5.04%
-   sd2: 0.3510682

```{r}
w <- 0.3483403
m1 <- -0.03484159
m2 <- 0.7023864
sd1 <- 0.6321144
sd2 <- 0.3510682
hist(data, 30, col="black", freq=F)
x <- seq(-3, 3, 0.1)
points(x, w*dnorm(x, m1, sd1)+(1-w)*dnorm(x,m2,sd2), "l", col="red", lwd=2)
```

```{r}
ploss %>% 
  group_by(year) %>% 
  sample_n(10^4) %>% 
  ggplot(aes(-loss, col = year)) +
  geom_density(bw = 0.1) +
  theme_bw() +
  scale_x_log10() +
  geom_vline(xintercept = c(10^m1, 10^m2)) +
  geom_vline(xintercept = 1, linetype = "dashed")
```

## Maps

### Area

```{r rastareadeforest}
#| eval: false
path <- "~/Documents/data/straore/"
deforest <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "deforest", full = TRUE) %>% 
  rast()
deforest_bin <- as.numeric(deforest < 0)
for(y in names(deforest_bin))
  deforest_bin[[y]] <- deforest_bin[[y]]*as.numeric(y)
deforest_bin[deforest_bin == 0] <- NA
deforest_tot <- min(deforest_bin, na.rm = T)
writeRaster(deforest_tot, file.path(path, "deforested_area.tif"))
```

```{r rastareadegrad}
#| eval: false
path <- "~/Documents/data/straore/"
degrad <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "floss", full = TRUE) %>% 
  rast()
degrad_bin <- as.numeric(degrad < 0)
for(y in names(degrad_bin))
  degrad_bin[[y]] <- degrad_bin[[y]]*as.numeric(y)
degrad_bin[degrad_bin == 0] <- NA
degrad_tot <- mean(degrad_bin, na.rm = T)
writeRaster(degrad_tot, file.path(path, "degraded_area.tif"))
```

```{r map}
path <- "~/Documents/data/straore/"
deforest <- rast(file.path(path, "deforested_area.tif"))
# degrad <- rast(file.path(path, "degraded_area.tif"))
ci <- geodata::gadm("Côte d'Ivoire", path = path, leval = 0) %>% 
  st_as_sf()
g1 <- ggplot(ci) +
  geom_sf(fill = NA) +
  tidyterra::geom_spatraster(data = deforest) +
  scale_fill_viridis_c("Year", na.value = NA) +
  theme_bw() +
  ggtitle("Deforestation")
# g2 <- ggplot(ci) +
#   geom_sf(fill = NA) +
#   tidyterra::geom_spatraster(data = degrad) +
#   scale_fill_viridis_c("Year", na.value = NA) +
#   theme_bw() +
#   ggtitle("Degradation")
# g1 + g2
g1
```

### Biomass

```{r}
#| eval: false
library(biscale)

deforest <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "deforest", full = TRUE) %>% 
  rast()
deforest_summary <- c(
  deforest[[as.character(2001:2007)]] %>% sum(),
  deforest[[as.character(2008:2015)]] %>% sum(),
  deforest[[as.character(2016:2022)]] %>% sum()
)
names(deforest_summary) <- c(2007, 2015, 2022)
deforest_summary[deforest_summary == 0] <- NA
degrad <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "floss", full = TRUE) %>% 
  rast()
degrad_summary <- c(
  degrad[[as.character(2001:2007)]] %>% sum(),
  degrad[[as.character(2008:2015)]] %>% sum(),
  degrad[[as.character(2016:2022)]] %>% sum()
)
names(degrad_summary) <- c(2007, 2015, 2022)
degrad_summary[degrad_summary == 0] <- NA

ci <- geodata::gadm("Côte d'Ivoire", path = path, leval = 0) %>% 
  st_as_sf() %>% 
  st_transform(crs = crs(deforest_summary)) %>% 
  as_Spatial() %>% 
  fortify() %>% 
  as.data.frame()
  
make_plot <- function(year) {
  stat <- as.data.frame(degrad_summary[[year]], xy = T) %>% 
    rename(degradation = year) %>% 
    full_join(as.data.frame(deforest_summary[[year]], xy = T)) %>% 
    rename(deforestation = year) %>% 
    mutate(degradation = -degradation/21.22, deforestation = -deforestation/21.22) %>% 
    mutate_all(~ ifelse(is.na(.), 0, .))
  data <- bi_class(stat, x = degradation, y = deforestation, style = "fisher", dim = 4)
  ggplot() +
    geom_polygon(data = ci, aes(x = long, y = lat,
                                 group = group), 
               fill = NA, col = "black") +
    geom_raster(data = data , aes(x = x, y = y, fill = bi_class)) +
    bi_scale_fill(pal = "PinkGrn", dim = 4) +
    bi_theme(base_size = 16) +
    theme_bw() +
    coord_sf() +
    theme(legend.position="none", axis.ticks = element_blank(),
        axis.title = element_blank(), axis.text = element_blank())
} 

stat <- as.data.frame(degrad_summary, xy = T) %>% 
  gather(year, degradation, -x, -y) %>% 
    full_join(as.data.frame(deforest_summary, xy = T) %>% 
                gather(year, deforestation, -x, -y)) %>% 
    mutate(degradation = -degradation/21.22, deforestation = -deforestation/21.22) %>% 
    mutate_all(~ ifelse(is.na(.), 0, .))

legend <- bi_legend(pal = "PinkGrn",
                    dim = 4,
                    xlab = "Degradation (tC/ha)",
                    ylab = "Deforestation (tC/ha)",
                    breaks = bi_class_breaks(stat, x = degradation, 
                                             y = deforestation, style = "fisher",
                dim = 4, dig_lab = c(x = 1, y = 1), split = T),
                    size = 10)

g <- cowplot::plot_grid(
  make_plot("2007"), make_plot("2015"), 
  make_plot("2022"), legend, label_y = 0.98,
  labels = c("2000-2007", "2008-2015", "2016-2022", "")
)

cowplot::save_plot(file.path(path, "comparison.png"), g, dpi = 1000, 
                   base_width = 10, base_height = 10, bg = "white")
```

```{r}
#| eval: false
library(colorplaner) # remotes::install_github("wmurphyrd/colorplaner")
ci <- geodata::gadm("Côte d'Ivoire", path = path, leval = 0) %>% 
  st_as_sf()
ci_df <- ci %>% 
  as_Spatial() %>% 
  fortify() %>% 
  as.data.frame()
deforest <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "deforest", full = TRUE) %>% 
  rast()
deforest_summary <- c(
  deforest[[as.character(2001:2007)]] %>% sum(),
  deforest[[as.character(2008:2015)]] %>% sum(),
  deforest[[as.character(2016:2022)]] %>% sum()
)
names(deforest_summary) <- c("2000-2007", "2008-2015", "2016-2022")
deforest_summary[deforest_summary == 0] <- NA
deforest_summary <- project(deforest_summary, crs(ci))
degrad <- list.files("~/Documents/data/straore/agb_change/", 
           pattern = "floss", full = TRUE) %>% 
  rast()
degrad_summary <- c(
  degrad[[as.character(2001:2007)]] %>% sum(),
  degrad[[as.character(2008:2015)]] %>% sum(),
  degrad[[as.character(2016:2022)]] %>% sum()
)
names(degrad_summary) <- c("2000-2007", "2008-2015", "2016-2022")
degrad_summary[degrad_summary == 0] <- NA
degrad_summary <- project(degrad_summary, crs(ci))
stat <- as.data.frame(degrad_summary, xy = T) %>% 
  gather(year, degradation, -x, -y) %>% 
    full_join(as.data.frame(deforest_summary, xy = T) %>% 
                gather(year, deforestation, -x, -y)) %>% 
    mutate(degradation = -degradation/21.22, deforestation = -deforestation/21.22)

stat2 <- stat %>% 
  mutate_all(~ ifelse(is.na(.), 0, .)) %>% 
  mutate(degradation = degradation+1, 
         deforestation=deforestation+1)

relab <- function(val) round(val-1)

g.comp <- ggplot() +
  geom_polygon(data = ci_df, aes(x = long, y = lat,
                                 group = group), 
               fill = NA, col = "black") +
  geom_tile(data = stat2, aes(x = x, y = y, 
                             fill = degradation, 
                             fill2 = deforestation)) +
  scale_fill_colorplane(name = "", 
                        axis_title = "Degradation",
                        axis_title_y = "Deforestation (tC/ha)",
                        trans = "log",
                        labels = relab,
                        labels_y = relab) +
  facet_wrap(~ year, nrow = 2) +
  theme_bw() +
  coord_sf() +
  theme(legend.position = c(0.8, 0.2), legend.key.size = unit(30, "pt")) +
  xlab("") + ylab("")
g.deforest <- ggplot() +
  geom_polygon(data = ci_df, aes(x = long, y = lat,
                                 group = group), 
               fill = NA, col = "black") +
  geom_tile(data = stat, 
            aes(x = x, y = y, fill = deforestation)) +
  scale_fill_viridis_c("Deforestation (tC/ha)", 
                       na.value = "grey90",
                       labels = round, direction = -1) +
  facet_wrap(~ year, nrow = 2) +
  theme_bw() +
  coord_sf() +
  theme(legend.position = c(0.8, 0.2)) +
  xlab("") + ylab("")
g.degrad <- ggplot() +
  geom_polygon(data = ci_df, aes(x = long, y = lat,
                                 group = group), 
               fill = NA, col = "black") +
  geom_tile(data = stat, aes(x = x, y = y, 
                             fill = degradation)) +
  scale_fill_viridis_c("Degradation (tC/ha)", na.value = "grey90",
                       labels = round, direction = -1) +
  facet_wrap(~ year, nrow = 2) +
  theme_bw() +
  coord_sf() +
  theme(legend.position = c(0.8, 0.2)) +
  xlab("") + ylab("")
ggsave(file.path(path, "deforestation.png"), g.deforest, dpi = 1000, 
       width = 10, height = 10, bg = "white")
ggsave(file.path(path, "degradation.png"), g.degrad, dpi = 1000, 
       width = 10, height = 10, bg = "white")
ggsave(file.path(path, "comparison.png"), g.comp, dpi = 1000, 
       width = 10, height = 10, bg = "white")
```

## Climatic zones

```{r climzoneci}
#| warning: false
ci <- st_read("~/Documents/data/limits/ci/ci.shp", quiet = TRUE)
cz <- st_read("~/Documents/data/CI/Zone_phytogeographique_rci/Zone_climatique.shp", 
              quiet = TRUE)%>% group_by(Zone) %>% 
  summarise() %>% 
  st_cast("POLYGON")
ggplot() +
  geom_sf(data = cz, aes(fill = Zone)) +
  geom_sf(data = ci, fill = NA) +
  theme_bw()
```

## WorldClim

One file is 8.5 Go for a resolution of 30 seconds against 33.4 Mo for a resolution of 10 minutes. Explorations will be done at 10 minutes.

We need:

-   2 30-years periods: 2030 & 2050
-   2 BioClimatic variables (BIO): BIO11 & BIO19
-   3 Scenarios (SSP): SSP 1 2.6, SSP 2 4.5, & SSP 5 8.5
-   22 Models (GCM): ACCESS-CM2, CMCC-ESM2, EC-Earth3-Veg, FIO-ESM-2-0, GISS-E2-1-G, HadGEM3-GC31-LL, INM-CM5-0, IPSL-CM6A-LR, MIROC6, MPI-ESM1-2-HR, MRI-ESM2-0, UKESM1-0-LL

> BIO11 = Mean Temperature of Coldest Quarter and BIO19 = Precipitation of Coldest Quarter

```{r dproj, eval=FALSE}
download_proj <- function(
    base_url = "https://geodata.ucdavis.edu/cmip6",
    resolution = "10m",
    model = "ACCESS-CM2",
    scenario = "ssp126",
    version = "wc2.1",
    variable = "bioc",
    period = "2021-2040"
){
  file <- paste0(version, "_", resolution, "_", variable, "_",
                 model, "_", scenario, "_", period, ".tif")
  url <- paste0(base_url, "/", resolution, "/", model, "/", scenario, "/", file)
  if(!file.exists(file.path("~/Documents/data/CI/wc_raw", file)))
    download.file(url, file.path("~/Documents/data/CI/wc_raw", file))
}

models <- c("ACCESS-CM2", "CMCC-ESM2", "EC-Earth3-Veg", 
            "FIO-ESM-2-0", "GISS-E2-1-G", "HadGEM3-GC31-LL",
            "INM-CM5-0", "IPSL-CM6A-LR", "MIROC6", "MPI-ESM1-2-HR", "MRI-ESM2-0", "UKESM1-0-LL")
scenario <- c("ssp126", "ssp585")
period <- c("2021-2040", "2041-2060")
lapply(models, function(m) 
  lapply(scenario, function(s)
    lapply(period, function(p)
      download_proj(model = m, scenario = s, period = p)
    )
  )
)

list.files("~/Documents/data/CI/wc_raw") %>% 
  lapply(function(file) {
    rast(file.path("~/Documents/data/CI/wc_raw", file)) %>% 
      crop(ci) %>% 
      select(bio11, bio19) %>% 
      project(crs(cz)) %>% 
      writeRaster(file.path("~/Documents/data/CI/wc", file),
                  overwrite = TRUE)
  })
```

```{r projs}
extract_proj <- function(file) {
  r <- rast(file.path("~/Documents/data/CI/wc", file))
  lapply(c("min", "mean", "max"), function(m)
    extract(r, cz, fun = m, na.rm=TRUE) %>% 
      mutate(zone = cz$Zone) %>% 
      select(-ID) %>% 
      gather(variable, value, -zone) %>% 
      mutate(metric = m)
  ) %>% bind_rows() %>% 
    mutate(file = file)
}

projs <- list.files("~/Documents/data/CI/wc") %>% 
  lapply(extract_proj) %>% 
  bind_rows() %>% 
  separate(file, c("version", "resolution", "var_type", "gcm", "ssp", "period"), "_") %>% 
  mutate(period = gsub(".tif", "", period)) %>% 
  pivot_wider(names_from = metric, values_from = value) %>% 
  mutate(zone = gsub("Secteur ", "", zone))
```

```{r bio11}
projs %>% 
  filter(variable == "bio11") %>% 
  ggplot(aes(gcm, mean)) +
  geom_point() +
  geom_linerange(aes(ymin = min, ymax = max)) +
  theme_bw() +
  facet_grid(zone ~ period + ssp) +
  ylab("Mean Temperature of Coldest Quarter (BIO11, °C)") + xlab("") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 5))
```

```{r bio19}
projs %>% 
  filter(variable == "bio19") %>% 
  ggplot(aes(gcm, mean/10^3)) +
  geom_point() +
  geom_linerange(aes(ymin = min/10^3, ymax = max/10^3)) +
  theme_bw() +
  facet_grid(zone ~ period + ssp) +
  ylab("Precipitation of Coldest Quarter (BIO19, m)") + xlab("") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 5), 
        axis.text.x = element_text(size = 8, angle = 45))
```

### Present versus future

```{r chull}
#| warning: false
future <- list.files("~/Documents/data/CI/wc/", full.names = T) %>% 
  lapply(rast) %>% 
  lapply(values) %>% 
  lapply(as.data.frame) %>% 
  bind_rows()
present <- data.frame(
  bio11 = rast("~/Documents/data/CI/wc2.1_10m_bio/wc2.1_10m_bio_11.tif") %>% 
  crop(ci) %>% 
  values() %>% 
  as.vector(),
  bio19 = rast("~/Documents/data/CI/wc2.1_10m_bio/wc2.1_10m_bio_19.tif") %>% 
  crop(ci) %>% 
  values() %>% 
  as.vector()
)
ggplot(future, aes(bio11, bio19)) +
  geom_point(alpha = 0.1, aes(col = "future")) +
  geom_point(alpha = 0.5, aes(col = "present"), data = present) +
  microViz::stat_chull(aes(col = "future")) +
  microViz::stat_chull(aes(col = "present"), data = present) +
  theme_bw() +
  xlab("Mean Temperature of Coldest Quarter (BIO11, °C)") + 
  ylab("Precipitation of Coldest Quarter (BIO19, mm)")
```

## 
